Rule learning for verification from temporal data

### 2016-08-31 Travis/secs/intro.tex
FMCAD - first draft

Machine learning is a new direction in verification\cite{Santolucito2016,ETH_Guy}.
Amoung the popular algorithms are neural nets and n-gram models, which produce to probabilistic models of correctness.
While often effective in practice, these do not provide the guarantees of a traditional verification approach.
In addition, these tools are not designed to provide simple justifications for their classification outputs.

Version space learning is a machine learning strategy for logical constraint classification\cite{mitchell82}. 
In addition to classification, version spaces have been used for programming-by-example \cite{lau2000version} and error detection \cite{Santolucito2016}.
This approach learns over a set of examples that have been classified into positive and negative examples.
We extend this approach to make use of classified examples with temporal structure, specifically by looking at the differences between sequential examples.

As a concrete use case, we plan to implement our algorithm to check for TravisCI configuration errors.
TravisCI is a continous integration tool that allows programmers to automatically run their test suite on every code update.
A recent usage study of TravisCI found that 15-20\% of failed TravisCI builds are due to "errors" - which means the configuration file was malformed and the software could not even be built \cite{API}.
Using the data from \cite{API}, we can also learn that since the start of 2014, approximately 88,000 hours of server time was used on TravisCI projects that resulted in an error status.
This number not only represents lost server time, but also lost developer time, as programmers must wait to verify that their work does not break the build.
If these malformed projects could be quickly statically checked on the client side, both TravisCI and its users could benefit.

>>>>

Machine learning is a powerful tool for software analysis and verification \cite{Santolucito2016,gehrpsi,garg2014ice}.
Among the popular algorithms are neural nets and n-gram models, which produce to probabilistic models of correctness.
While often effective in practice, these do not provide the guarantees of a traditional verification approach.
In addition, these tools are not designed to provide simple justifications for their classification outputs.

In contrast, version space learning is a machine learning strategy for logical constraint classification \cite{mitchell82}, appropriate when the learning set does not contain noise.
Extensions to version space learning have been used various software analysis techniques, such as programming-by-example \cite{lau2000version}, invariant synthesis \cite{garg2014ice}, and error detection \cite{Santolucito2016}.
We first give a formal definition of the algorithm in \cite{Santolucito2016} in terms of version space learning.
We then extend that work to leverage temporal structure in a learning set, specifically by looking at the differences between sequential examples.

As a concrete use case, we plan to implement our algorithm to check for TravisCI configuration errors.
TravisCI is a continuous integration tool that allows programmers to automatically run their test suite on every code update.
A recent usage study of TravisCI found that 15-20\% of failed TravisCI builds are due to "errors" - which means the configuration file was malformed and the software could not even be built \cite{API}.
Using the data from \cite{API}, we can also learn that since the start of 2014, approximately 88,000 hours of server time was used on TravisCI projects that resulted in an error status.
This number not only represents lost server time, but also lost developer time, as programmers must wait to verify that their work does not break the build.
If these malformed projects could be quickly statically checked on the client side, both TravisCI and its users could benefit.

### 2016-08-31 Travis/secs/temporal.tex
FMCAD - first draft

An interesting extension to ConfigC on a theoretical side is to look at the temporal relations between commits.
The key observation is that with each commit that changes the build status, we can learn highly localized information about our model.

The first step is to build an intermediate representation of the data we will learn.
This data must be structured as a shallow tree for generalizable learning.
This restriction is why this approach is not appropriate for language learning on large grammars (such as a programming language).

While the travis.yml file has a simple structure and can be taken verbatim, we must use a simplification of the program code.
Since we are only interested in the parts of the program effecting build status, we should extract key feautres such as programming language and a list of imported libraries.
We will call this program summary $P_t$, a simplified representation of the repository which contains only the information relevant to the learning process.
The summary must contains every piece of information that might lead to a build error.

The subscript on $P_t$ is a tag based on the ordered commit history.
However, a git history is not a limited to a single linear timeline.
Git features the ability to \textit{branch}, which allows to simultanious commit chains.
To handle the start of a branch, add a superscript to indicate the branch, and restart the counter on a branch.
To handle the merge of two branches $P_{t}^{x}$ and $P_{t'}^{y}$, step to $P_{t+1}^{x}$, where $x$ is the mainline branch.
We then say that $P_{t'}^{y}$ has no successor commit $P_{t'+1}^{y}$.

From this summary we can then build a model $M(P_t)$, which is the full set of possible rules derivable from the program summary.
These possible rules must be given by the user.
In \cite{Santolucito2016}, these were called Attributes, and express constraints such as, ordering of lines, or substring relations on values.

We will denote the build status of $P_t$ with $B(P_t)$.
In this application, we consider only the passing and erroring build status, denoted $Pass$, and $Err$ respectively.
All status that are not \verb|error|, as defined by the Travis API, will be included as passing.
For brevity, we denote sequences of build statuses with the following notation:

@@

With this notation, we can formally express the requirement that the program summary is complete.
This is an assumption on the quality of the program summary.

\begin{align*}
  \forall S(P_t)=Err, \exists r \in M(P_t), r \in Br
\end{align*}

From the above we know that if a build is erroring, then there must exist at least one error.
By pushing the negation into the formula, we can also know that if a build is passing, then there must not exist any errors. 
That is, the model of a passing commit must not contain any rules which are breaking. 
Note we are not, however, garunteed that any rules from a passing commit are necessary.

@@

While Eq. \ref{eq:E} and \ref{eq:P} might build a basic model, they will do not capture all of the available knowledge.
The key insight is that when we commit a break (P E), we can localize the error to one of the lines that changed.
Either we removed something that was necessary, or added something that was breaking. 
We use an inclusive disjunction, since a erroring commit can break multiple things at once. 
Expressed formally, where (\ is the set difference), that is:

@@

\subsection{Unsatisfiability}
Note we have made a strong assumption that the summary $P_t$ contains every piece of information that might lead to a build error. 
This is the strongest assumption that we make - however our algorithm is able to detect cases where the summary is incomplete.
If we cannot find a solution for Eq. 3 it means that the set of rules we are looking for is incomplete.

>>>>

%The first step is to build an intermediate representation of the data we will learn.
%This data must be structured as a shallow tree for generalizable learning.
%This restriction is why this approach is not appropriate for language learning on large grammars (such as a programming language).

In ConfigC, the configuration files were analyzed as standalone documents.
Since a TravisCI configuration file is dependent on the code it is trying to build, we must consider a more general sense of configuration file.
We will call this a program summary $P_t$, which is a representation of the repository which contains the information relevant to the learning process.
In the case of TravisCI, this include the \verb|.travis.yml| file, as well as extract key code features that may effect build status, such as programming language and a list of imported libraries.
The summary must contain every piece of information that might lead to a build error.

The subscript on $P_t$ is a time stamp tag based on the ordered commit history.
However, a git history is not a limited to a single linear timeline.
Git features the ability to \textit{branch}, which allows to simultaneous commit chains.
To handle the start of a branch, add a superscript to indicate the branch, and restart the counter on a branch.
To handle the merge of two branches $P_{t}^{x}$ and $P_{t'}^{y}$, step to $P_{t+1}^{x}$, where $x$ is the mainline branch.
We then say that $P_{t'}^{y}$ has no successor commit $P_{t'+1}^{y}$.

We will denote the build status of $P_t$ with $S(P_t)$.
In this application, we consider only the passing and erroring build status, denoted $Pass$, and $Err$ respectively.
All status that are not \verb|error|, as defined by the Travis API, will be included as passing.
For brevity, we denote sequences of build statuses with the following notation:

@@

From this summary we can then build a model $M(P_t)$, as in ConfigC, which is the full set of possible relations derivable from the program summary.
In contrast with ConfigC, we now consider both positive and negative examples and so must introduce the \textit{general boundary}.
The general boundary is the dual of the specific boundary, and is the most relaxed requirement for a positive classification.
We denoted specific boundary as the set of necessary relations $Nec$, and now denote the general boundary as the set of breaking relations $Br$.
With this notation, we can formally express the requirement that the program summary is complete.

\begin{align}
  \forall S(P_t)=Err, \exists r \in M(P_t), r \in Br \label{eq:E1}
\end{align}

From the above we know that if a build is erroring, then there must exist at least one error.
By pushing the negation into the formula, we can also know that if a build is passing, then there must not exist any errors.
That is, the model of a passing commit must not contain any rules which are breaking.
Note we are not, however, guaranteed that any rules from a passing commit are necessary.

@@

While Eq. \ref{eq:E} and \ref{eq:P} might build a basic model, they will do not capture all of the available knowledge.
The key insight is that when we commit a break (P E), we can localize the error to one of the lines that changed.
Either we removed something that was necessary, or added something that was breaking.
We use an inclusive disjunction, since a erroring commit can break multiple things at once.
Expressed formally, where $\setminus$ is the set difference, that is:

@@
